Male-biased sexual selection, but not sexual dichromatism, predicts speciation in birds

### 2020-08-23 Manuscript_Publication.Rmd
slight changes to manuscript and figure

---
title: Male-biased sexual selection, but not sexual dichromatism, predicts speciation in birds
author:
bibliography: References/SS_Extinction.bib
output:
  pdf_document:
    keep_tex: yes
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[L]{\textsl{\leftmark}}
  \rhead[]{Sexual selection predicts speciation} \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
  \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption}
  \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
  \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip}
  \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape}
  \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{}
  \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{}
  \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} \usepackage[hang,
  flushmargin]{footmisc} \usepackage{hyperref}\hypersetup{draft, colorlinks=false, hyperfootnotes=false,
  linkcolor=blue, filecolor=blue, urlcolor=blue} \usepackage{xcolor} \renewcommand\thefootnote{\textcolor{black}{\arabic{footnote}}}
  \raggedbottom
csl: References/evolution.csl
subparagraph: yes
abstract: Sexual selection is thought to shape phylogenetic diversity by affecting speciation or extinction rates. However, the net effect of sexual selection on diversification is hard to predict _a priori_, because many of the hypothesised effects on speciation or extinction have opposing signs and uncertain magnitudes. **Theoretical work also suggests that the net effect of sexual selection on diversification should depend strongly on ecological factors, though this prediction has seldom been tested. Here, we test whether variation in sexual selection can predict speciation and extinction rates across passerine birds (up to 5,812 species, covering most genera) and whether this relationship is mediated by environmental factors. Male-biased sexual selection, and specifically sexual size dimorphism, predicted two of the three measures of speciation rates that we examined**. The link we observed between sexual selection and speciation was independent of environmental variability, though species with smaller ranges had higher speciation rates. There was no association between sexual selection and extinction rate. **Our findings support the view that male-biased sexual selection, as measured by predictors of male-male competition, has shaped diversification in the largest radiation of birds.**
---
<!-- \maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Parkville, VIC, 3052, Australia} -->

@@

Birds have been a popular focus of macroevolutionary studies of sexual selection and diversification [@Barraclough_1995; @Morrow_Speciation_2003; @Seddon_2008; @Seddon_2013; @Huang_2014] because their biology and phylogenetic relationships are comparatively well-known. A 2011 meta-analysis covering 20 primary studies of birds and other taxa found a small but significant positive association between sexual selection and speciation, with the average effect size in birds stronger than in mammals but weaker than in insects or fish [@Kraaijeveld_2011]. However, there was large variation in effect size estimates across the 20 studies, likely reflecting differences in methodology, such as metrics used to characterise speciation and sexual selection, in addition to true biological differences. More recently, @Huang_2014 found no association between sexual dichromatism and speciation (*n* = 918 species) in a study using spectrophotometric measurements of museum specimens [@Armenta_2008] and tip-rate estimates from a molecular-only phylogeny [@Jetz_2012]. Similarly, @Cooney_2017 found no effect of sexual dichromatism on diversification across 1,306 pairs of species, using dichromatism scores provided by human observers.More recently, social polygyny (a proxy for sexual selection) was found to have a positive association with speciation rate across 954 species of birds [@Iglesias-Carrasco_2019].

Here, we investigate the association between sexual selection and diversification in birds while building upon previous approaches in multiple ways. We use two measures of the strength of sexual selection: sexual dichromatism [@Dale_2015], as well as an index of male-biased sexual selection [@Dale_2015], which captures (co)variation in sexual size dimorphism, social polygyny and paternal care. We use these two measures because sexual dichromatism does not always signal the presence of strong sexual selection and *vice versa* [@Dale_2015]. For example, male and female dunnocks (*Prunella modularis*) are similarly coloured yet sexual selection appears to be strong [@davies1986re]. Furthermore, a recent comparative study found a negative relationship between dichromatism and another sexually-selected trait (song) across species, suggesting that a multi-trait focus would improve estimates of sexual selection intensity [@Cooney_2018]. Additionally, our analysis includes multiple ecological and environmental variables, allowing us to adjust for potential confounds, to identify environmental factors driving diversification, and to test whether environmental factors interact with sexual selection as theory predicts [@Connallon_Hall_2016]. We use multiple approaches for quantifying speciation and extinction rates at the tips of phylogenetic trees, including Bayesian Analysis of Macroevolutionary Mixtures [BAMM; @Beaulieu_2015; @Rabosky_Extinction_2016; @Moore_2016; @Rabosky_BAMM_flawed_2017], as well as older but reliable tip-rate statistics, namely diversification rate ($\lambda_{DR}$) and node density ($\lambda_{ND}$) [@Jetz_2012]. Our results show that (*i*) a composite measure of sexual selection, but not sexual dichromatism, significantly predicts speciation rates, (*ii*) species with smaller ranges have higher speciation rates and (*iii*) there is no evidence that environmental variables or their interaction with sexual selection have an impact on diversification rates. Therefore, we provide evidence at a very large scale that sexual selection can have positive effects on diversification in the largest radiation of birds. Furthermore, we suggest that the use of sexual dichromatism as the sole proxy for sexual selection should be reconsidered, since it appears to be inconsistently associated with the operation of sexual selection.

@@

Secondly, we used BAMM to model the dynamics of speciation and extinction across the **101 phylogenetic trees (one MCC tree and 100 random draws of the posterior)**. This software uses a Bayesian approach (reversible-jump Markov Chain Monte Carlo) to generate probability distributions of evolutionary rate-shift configurations with variable speciation and extinction rates [@Rabosky_2014]. **These models provide tip-rate estimates of speciation and extinction rate that can be easily used in comparative analyses**. The parameters of the 100 BAMM runs are detailed in full in the \href{https://justincally.github.io/SexualSelection_Speciation/#set_up_bamm_parameters}{Supplementary Information}; **briefly, we used a time-variable model with the prior expected number of evolutionary rate shifts set at 100 and prior rates set from the initial speciation and extinction values using the ``BAMMtools`` R package [@bammtools_2014]. BAMM models were run independently for the 101 phylogenetic trees for 100 million generations, and given the computationally intensive nature of BAMM, runs were conducted across multiple CPUs. Important BAMM parameters (log-likelihood and number of rate shifts) reached convergence with effective sample size (ESS) of MCMC (Markov Chain Monte Carlo) samples surpassing 200; an arbitrary value, above which posterior distributions can often be accurately inferred** (\href{https://justincally.github.io/SexualSelection_Speciation/#bamm_measures_of_speciation_and_extinction}{Table S3}, \href{https://justincally.github.io/SexualSelection_Speciation/#bamm_measures_of_speciation_and_extinction}{Table S4}). Further details of BAMM parameters and output are available in the \href{https://justincally.github.io/SexualSelection_Speciation/}{Supplementary Information}, with tip-rate means and variances provided. Additionally, given the variability in BAMM estimates, we also provide analysis of BAMM shift configurations and tip-rate estimates from our run on the MCC tree and within a BAMM run on the MCC tree from a genetic-only phylogeny across all birds [@Harvey_2017]. All analyses were conducted on log-rates.  \phantomsection

@@

To test the association between speciation/extinction and sexual selection, environmental variability and their interaction, we used phylogenetic least squares (PGLS) models in the ``nlme`` package [@nlme_2018]. Firstly, we conducted model selection to compare models in which $\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$ or $\mu_{BAMM}$ were the response variable: these tip-rate estimates all came from the same MCC tree (derived from 2,500 draws of the posterior distribution [@Jetz_2012]). For models of $\lambda_{BAMM}$ and $\mu_{BAMM}$ we used the inverse of the variance associated with each tip rate estimate as weights, to account for the variable precision of the estimates provided by BAMM. **For each response variable, we conducted model selection to compare models with different combinations of predictor variables. ** The most complicated model in each set under comparison contained one of the measures of sexual selection (sexual dichromatism or the index of male-biased sexual selection), all of the environmental measures (i.e. log-transformed range size, seasonal temperature variation, spatial temperature variation, long-term temperature variation, and NPP), and all of the 2-way interactions between sexual selection and each of the environmental measures. The simpler models contained all of the same main effects, but had fewer 2-way interaction terms (potentially none). Model selection was done in ``MuMIn`` using the ``dredge`` function [@MuMIn_2017]. Using the terms from the top-ranked model (ranked by AICc), we ran the equivalent model for each of the 1,000 phylogenetic trees used to derive $\lambda_{DR}$, $\lambda_{ND}$ and each of the 100 trees used to derive $\lambda_{BAMM}$ and $\mu_{BAMM}$. **Additionally, we investigated the effect of the individual principal component scores for the index of male-biased sexual selection on speciation rate. For these pgls models we replaced the composite index score with the individual biological variable (sexual size dimorphism, social polygyny and [lack of] paternal care) and ran the equivalent model for 300 phylogenetic trees used to derive $\lambda_{DR}$, $\lambda_{ND}$ and 100 trees used to derive $\lambda_{BAMM}$**.

@@

Finally, using the subset of species with an index of male-biased sexual selection, we conducted a phylogenetic path analysis using the ``phylopath R`` package [@Wouter_2018]. The phylogenetic path analysis was used to assess causal paths between variables not able to be modelled within the univariate response of PGLS. That is, a phylogenetic path analysis allowed us to model relationships between the predictor variables used in our PGLS analysis as we anticipate environmental variability, sexual dichromatism/selection, and range size to have effects on each other and not just on speciation rate. To minimise path complexity we used temperature seasonality (BIO4) as the single measure for environmental variability, $\lambda_{DR}$ as the single measure of speciation and the tip-rates from the MCC tree. Further details of the path analysis, including our rationale for each path's directions, can be found within the \href{https://justincally.github.io/SexualSelection_Speciation/}{Supplementary Information} along with all other analyses and the relevant ``R`` code to reproduce results.  \phantomsection

@@

**Given that this male-biased sexual selection index comprises three different (but correlated) variables, we investigated which of these variables was driving the association observed with $\lambda_{DR}$. Our results over 300 trees showed that this pattern is mainly driven by the sexual size dimorphism component (HPD Interval = `r sanitize.numbers(formatC(dr.wing, format = "e", digits = 2, drop0trailing = T), type = "latex", math.style.exponents = T)`), with the effects of other components overlapping zero; paternal care (HPD Interval = `r sanitize.numbers(formatC(dr.paternal, format = "e", digits = 2, drop0trailing = T), type = "latex", math.style.exponents = T)`) and mating system (HPD Interval= `r sanitize.numbers(formatC(dr.mating, format = "e", digits = 2, drop0trailing = T), type = "latex", math.style.exponents = T)`). Importantly, the association between sexual size dimorphism and speciation rates is also present when using $\lambda_{ND}$ (HPD Interval = `r sanitize.numbers(formatC(nd.wing, format = "e", digits = 2, drop0trailing = T), type = "latex", math.style.exponents = T)`), but not when using $\lambda_{BAMM}$ (HPD Interval = `r sanitize.numbers(formatC(b.wing, format = "e", digits = 2, drop0trailing = T), type = "latex", math.style.exponents = T)`, \autoref{SS_components}).**

@@

Estimates of the effect of predictor variables on speciation rates varied across phylogenetic trees, especially in the BAMM rates ($\lambda_{BAMM}$ and $\mu_{BAMM}$), where the 95 % HPD interval across PGLS model estimates from 100 trees was often > 20 times larger than the 95 % confidence interval for estimates from a single PGLS model using the MCC tree. This contrasts with variation across trees for the other rate estimates ($\lambda_{DR}$ and $\lambda_{ND}$), where the 95 % HPD interval of model estimates for pgls models using 1,000 trees was near-equivelant to the 95 % confidence interval calculated for PGLS model estimates of the MCC tree (\href{https://justincally.github.io/SexualSelection_Speciation/#analysis}{Table S9}). Given the computational requirements of BAMM, the great majority of earlier studies have based their estimates on a single consensus tree, **and so it is concerning that there are such large differences in the BAMM estimates between alternative, similarly plausible phylogenies**. Mean measures of speciation rate across 100 trees were positively correlated between measures ($\lambda_{DR}$ - $\lambda_{BAMM}$: *r*=`r round(cor(all.trees.join$DR, all.trees.join$mean.lambda), digits = 2)`, $\lambda_{DR}$ - $\lambda_{ND}$: *r*=`r round(cor(all.trees.join$DR, all.trees.join$ND), digits = 2)`, $\lambda_{ND}$ - $\lambda_{BAMM}$: *r*=`r round(cor(all.trees.join$ND, all.trees.join$mean.lambda), digits = 2)`; \href{https://justincally.github.io/SexualSelection_Speciation/#additional_figures_and_tables}{Figure S15}). Given that the calculation of BAMM rates can be affected by the settings of the run and the use of different priors, we compared the estimate of our MCC tree with that of previous published analyses on birds, and found a high correlation (*r*=`r round(cor(Harvey.vs.us$mean.lambda, Harvey.vs.us$Harvey.lambda, use = "pairwise"), digits = 2)`, \href{https://justincally.github.io/SexualSelection_Speciation/#bamm_measures_of_speciation_and_extinction}{Figure S8}). Full details of the BAMM results are presented as supplementary materials.

@@

![Estimates of the effect of individual sexual selection components included in the PPCA (paternal care and wing size dimorphism) on three measures of speciation rate ($\lambda_{DR}$, $\lambda_{ND}$ and $\lambda_{BAMM}$). Estimates are presented as density intervals from pgls models on 300 phylogentic trees that used species with available data for these sexual selection measures (*n* = 2,465). The bar under each density ridge is the 95 % Highest Posterior Density Interval. \label{SS_components}](Figures/Individual_SS_Estimates.pdf)

@@

**Sexual selection due to competition within sexes contrasts with the drivers of sexual dichromatism.  Plumage dichromatism can evolve as a consequence of female cryptic choice and be related to extra-bond fertilizations, but not necessarily paternal care or mating system [@owens_1998]. It can also arise as a result of selection on the level of crypsis of the sex that cares for offspring [@Dale_2015]. The fact that traits linked with competition (such as size dimorphism) are the ones associated with higher $\lambda_{DR}$ values – rather than sexual dichromatism – supports the general view that antagonistic interactions and sexual conflict can lead to increased diversity [@Bonduriansky_2011; @Hoskin_2010; @tsuji2020sexual]. Moreover, body size is a trait that influences multiple aspects of the physiology and ecology of a species. Differences in body size (as a result of sexual selection) could be linked to changes in diet, vulnerability to predators or environmental tolerance [@Damuth_1993; @Liow_2008; @Bonduriansky_2011], and such differences could ultimately increase the likelihood of divergence between young lineages. In mammals, sexual selection has driven the evolution of large body size which in turn has allowed diversification of ecological strategies in the clade, and higher speciation rates.**

@@

In addition to speciation, sexual selection is hypothesised to affect extinction. Using the model-based approach of BAMM, we found no association between the estiamted extinction rate and sexual dichromatism, male-biased sexual selection, or our measures of environmental variability. However, these extinction results should not be regarded as definitive because extinction is notoriously difficult to estimate accurately from phylogenies, principally because different combinations of speciation and extinction rates can give rise to similar patterns of diversity [see @Rabosky_Extinction_2016]. Phylogenetic methods such as BAMM allow for speciation and extinction rates to be estimated using moderately-sized phylogenies, although the ability of BAMM to model evolutionary rate shifts and extinction rates is debated [see, @Beaulieu_2015; @Rabosky_Extinction_2016; @Moore_2016; @Rabosky_BAMM_flawed_2017]. Additionally, while several tip-rate estimates exist for speciation rate (e.g., $\lambda_{DR}$ and $\lambda_{ND}$), tip-rate estimates of extinction rate are difficult to obtain without complex Bayesian models which are sensitive to sampling bias [@Davis_2013]. Although extinction rates can be inferred from alternative sources, such as the fossil record [@Martins_2018], direct observation extinction, or IUCN red list status [@Greenberg_2017], each approach has its limitations. Across the passerine bird phylogeny, we found that BAMM often produced homogeneous speciation and extinction rates for smaller clades showing few rate shifts, which might reduce our power to detect small differences in extinction rates among closely-related taxa [@Rabosky_BAMM_flawed_2017; @Rabosky_diversification_2018]. Thus, this methodological constraint likely decreases our ability to accurately measure the correlation between metrics of sexual selection and the probability of extinction.

>>>>

---
title: Male-biased sexual selection, but not sexual dichromatism, predicts speciation in birds
author:
bibliography: References/SS_Extinction.bib
output:
  pdf_document:
    keep_tex: yes
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[L]{\textsl{\leftmark}}
  \rhead[]{Sexual selection predicts speciation} \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
  \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption}
  \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
  \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip}
  \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape}
  \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{}
  \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{}
  \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} \usepackage[hang,
  flushmargin]{footmisc} \usepackage{hyperref}\hypersetup{draft, colorlinks=false, hyperfootnotes=false,
  linkcolor=blue, filecolor=blue, urlcolor=blue} \usepackage{xcolor} \renewcommand\thefootnote{\textcolor{black}{\arabic{footnote}}}
  \raggedbottom
csl: References/evolution.csl
subparagraph: yes
abstract: Sexual selection is thought to shape phylogenetic diversity by affecting speciation or extinction rates. However, the net effect of sexual selection on diversification is hard to predict _a priori_, because many of the hypothesised effects on speciation or extinction have opposing signs and uncertain magnitudes. **Theoretical work also suggests that the net effect of sexual selection on diversification should depend strongly on ecological factors, though this prediction has seldom been tested. Here, we test whether variation in sexual selection can predict speciation and extinction rates across passerine birds (up to 5,812 species, covering most genera) and whether this relationship is mediated by environmental factors. Male-biased sexual selection, and specifically sexual size dimorphism, predicted two of the three measures of speciation rates that we examined**. The link we observed between sexual selection and speciation was independent of environmental variability, though species with smaller ranges had higher speciation rates. There was no association between any proxies of sexual selection and extinction rate. **Our findings support the view that male-biased sexual selection, as measured by predictors of male-male competition, has shaped diversification in the largest radiation of birds.**
---
<!-- \maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Parkville, VIC, 3052, Australia} -->

@@

Birds have been a popular focus of macroevolutionary studies of sexual selection and diversification [@Barraclough_1995; @Morrow_Speciation_2003; @Seddon_2008; @Seddon_2013; @Huang_2014] because their biology and phylogenetic relationships are comparatively well-known. A 2011 meta-analysis covering 20 primary studies of birds and other taxa found a small but significant positive association between sexual selection and speciation, with the average effect size in birds stronger than in mammals but weaker than in insects or fish [@Kraaijeveld_2011]. However, there was large variation in effect size estimates across the 20 studies, likely reflecting differences in methodology, such as metrics used to characterise speciation and sexual selection, in addition to true biological differences. More recently, @Huang_2014 found no association between sexual dichromatism and speciation (*n* = 918 species) in a study using spectrophotometric measurements of museum specimens [@Armenta_2008] and tip-rate estimates from a molecular-only phylogeny [@Jetz_2012]. Similarly, @Cooney_2017 found no effect of sexual dichromatism on diversification across 1,306 pairs of species, using dichromatism scores provided by human observers. More recently, social polygyny (a proxy for sexual selection) was found to have a positive association with speciation rate across 954 species of birds [@Iglesias-Carrasco_2019].

Here, we investigate the association between sexual selection and diversification in birds while building upon previous approaches in multiple ways. We use two measures of the strength of sexual selection: sexual dichromatism [@Dale_2015], as well as an index of male-biased sexual selection [@Dale_2015], which captures (co)variation in sexual size dimorphism, social polygyny and paternal care. We use these two measures because sexual dichromatism does not always signal the presence of strong sexual selection and *vice versa* [@Dale_2015]. For example, male and female dunnocks (*Prunella modularis*) are similarly coloured yet sexual selection appears to be strong [@davies1986re]. Furthermore, a recent comparative study found a negative relationship between dichromatism and another sexually-selected trait (song) across species, suggesting that a multi-trait focus would improve estimates of sexual selection intensity [@Cooney_2018]. Additionally, our analysis includes multiple ecological and environmental variables, allowing us to control for potential confounds, to identify environmental factors driving diversification, and to test whether environmental factors interact with sexual selection as theory predicts [@Connallon_Hall_2016].

We use multiple approaches for quantifying speciation and extinction rates at the tips of phylogenetic trees, including Bayesian Analysis of Macroevolutionary Mixtures [BAMM; @Beaulieu_2015; @Rabosky_Extinction_2016; @Moore_2016; @Rabosky_BAMM_flawed_2017], as well as older but reliable tip-rate statistics, namely diversification rate ($\lambda_{DR}$) and node density ($\lambda_{ND}$) [@Jetz_2012]. Our results show that (*i*) a composite measure of sexual selection, but not sexual dichromatism, significantly predicts speciation rates, (*ii*) **the significant association between the composite measure of sexual selection and speciation rate is largely driven by sexual size dimorphism**, (*iii*) species with smaller ranges have higher speciation rates and (*iv*) there is no evidence that environmental variables or their interaction with sexual selection have an impact on diversification rates. Therefore, we provide evidence at a very large scale that sexual selection can have positive effects on diversification in the largest radiation of birds. Furthermore, we suggest that the use of sexual dichromatism as the sole proxy for sexual selection should be reconsidered, since it appears to be inconsistently associated with the operation of sexual selection.

@@

We used BAMM to model the dynamics of speciation and extinction across the **101 phylogenetic trees (one MCC tree and 100 random draws of the posterior)**. This software uses a Bayesian approach (reversible-jump Markov Chain Monte Carlo) to generate probability distributions of evolutionary rate-shift configurations with variable speciation and extinction rates [@Rabosky_2014]. **These models provide tip-rate estimates of speciation and extinction rate that can be easily used in comparative analyses**. The parameters of the 100 BAMM runs are detailed in full in the \href{https://justincally.github.io/SexualSelection_Speciation/#set_up_bamm_parameters}{Supplementary Information}; **briefly, we used a time-variable model with the prior expected number of evolutionary rate shifts set at 100 and prior rates set from the initial speciation and extinction values using the ``BAMMtools`` R package [@bammtools_2014]. BAMM models were run independently for the 101 phylogenetic trees for 100 million generations, and given the computationally intensive nature of BAMM, runs were conducted across multiple CPUs. Important BAMM parameters (log-likelihood and number of rate shifts) reached convergence with effective sample size (ESS) of MCMC (Markov Chain Monte Carlo) samples surpassing 200; an arbitrary value, above which posterior distributions can often be accurately inferred** (\href{https://justincally.github.io/SexualSelection_Speciation/#bamm_measures_of_speciation_and_extinction}{Table S3}, \href{https://justincally.github.io/SexualSelection_Speciation/#bamm_measures_of_speciation_and_extinction}{Table S4}). Further details of BAMM parameters and output are available in the \href{https://justincally.github.io/SexualSelection_Speciation/}{Supplementary Information}, with tip-rate means and variances provided. Additionally, given the variability in BAMM estimates, we also provide analysis of BAMM shift configurations and tip-rate estimates from our run on the MCC tree and within a BAMM run on the MCC tree from a genetic-only phylogeny across all birds [@Harvey_2017]. All analyses were conducted on log-rates.  \phantomsection

@@

To test the association between speciation/extinction and sexual selection, environmental variability and their interaction, we used phylogenetic least squares (PGLS) models in the ``nlme`` package [@nlme_2018]. Firstly, we conducted model selection to compare models in which $\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$ or $\mu_{BAMM}$ were the response variable: these tip-rate estimates all came from the same MCC tree (derived from 2,500 draws of the posterior distribution [@Jetz_2012]). For models of $\lambda_{BAMM}$ and $\mu_{BAMM}$ we used the inverse of the variance associated with each tip rate estimate as weights, to account for the variable precision of the estimates provided by BAMM. **For each response variable, we conducted model selection to compare models with different combinations of predictor variables. ** The most complex model in each set under comparison contained one of the measures of sexual selection (sexual dichromatism or the index of male-biased sexual selection), all of the environmental measures (i.e. log-transformed range size, seasonal temperature variation, spatial temperature variation, long-term temperature variation, and NPP), and all of the 2-way interactions between sexual selection and each of the environmental measures. The simpler models contained all of the same main effects, but had fewer 2-way interaction terms (potentially none). Model selection was done in ``MuMIn`` using the ``dredge`` function [@MuMIn_2017]. Using the terms from the top-ranked model (ranked by AICc), we ran the equivalent model for each of the 1,000 phylogenetic trees used to derive $\lambda_{DR}$, $\lambda_{ND}$ and each of the 100 trees used to derive $\lambda_{BAMM}$ and $\mu_{BAMM}$. **Additionally, we investigated the effect of the individual variables used to derive the index of male-biased sexual selection on speciation rate. For these pgls models we replaced the composite index score with the individual biological variable (sexual size dimorphism, social polygyny and [lack of] paternal care) and ran the equivalent model for 300 phylogenetic trees used to derive $\lambda_{DR}$, $\lambda_{ND}$ and 100 trees used to derive $\lambda_{BAMM}$**.

@@

Finally, using the subset of species with an index of male-biased sexual selection, we conducted a phylogenetic path analysis using the ``phylopath R`` package [@Wouter_2018]. The phylogenetic path analysis was used to assess causal paths between variables unable to be modelled within the univariate response of PGLS. That is, a phylogenetic path analysis allowed us to model relationships between the predictor variables used in our PGLS analysis as we anticipate environmental variability, sexual dichromatism/selection, and range size to have effects on each other and not just on speciation rate. To minimise path complexity we used temperature seasonality (BIO4) as the single measure for environmental variability, $\lambda_{DR}$ as the single measure of speciation and the tip-rates from the MCC tree. Further details of the path analysis, including our rationale for each path's directions, can be found within the \href{https://justincally.github.io/SexualSelection_Speciation/}{Supplementary Information} along with all other analyses and the relevant ``R`` code to reproduce results.  \phantomsection

@@

**Given that this male-biased sexual selection index comprises of three different (but correlated) variables, we investigated which of these variables was driving the association observed with $\lambda_{DR}$. Our results over 300 trees showed that this pattern is mainly driven by the sexual size dimorphism component (HPD Interval = `r sanitize.numbers(formatC(dr.wing, format = "e", digits = 2, drop0trailing = T), type = "latex", math.style.exponents = T)`), with the effects of other components overlapping zero; paternal care (HPD Interval = `r sanitize.numbers(formatC(dr.paternal, format = "e", digits = 2, drop0trailing = T), type = "latex", math.style.exponents = T)`) and mating system (HPD Interval= `r sanitize.numbers(formatC(dr.mating, format = "e", digits = 2, drop0trailing = T), type = "latex", math.style.exponents = T)`). Importantly, the association between sexual size dimorphism and speciation rates is also present when using $\lambda_{ND}$ (HPD Interval = `r sanitize.numbers(formatC(nd.wing, format = "e", digits = 2, drop0trailing = T), type = "latex", math.style.exponents = T)`), but not when using $\lambda_{BAMM}$ (HPD Interval = `r sanitize.numbers(formatC(b.wing, format = "e", digits = 2, drop0trailing = T), type = "latex", math.style.exponents = T)`, \autoref{SS_components}).**

@@

Estimates of the effect of predictor variables on speciation rates varied across phylogenetic trees, especially in the BAMM rates ($\lambda_{BAMM}$ and $\mu_{BAMM}$), where the 95 % HPD interval across PGLS model estimates from 100 trees was often > 20 times larger than the 95 % confidence interval for estimates from a single PGLS model using the MCC tree. This contrasts with variation across trees for the other rate estimates ($\lambda_{DR}$ and $\lambda_{ND}$), where the 95 % HPD interval of model estimates for pgls models using 1,000 trees was near-equivelant to the 95 % confidence interval calculated for PGLS model estimates of the MCC tree (\href{https://justincally.github.io/SexualSelection_Speciation/#analysis}{Table S9}). Given the computational requirements of BAMM, the great majority of earlier studies have based their estimates on a single consensus tree, **and so it is concerning that there are such large differences in the BAMM estimates between alternative, similarly plausible phylogenies**. Mean measures of speciation rate across 100 trees were positively correlated between measures ($\lambda_{DR}$ - $\lambda_{BAMM}$: *r*=`r round(cor(all.trees.join$DR, all.trees.join$mean.lambda), digits = 2)`, $\lambda_{DR}$ - $\lambda_{ND}$: *r*=`r round(cor(all.trees.join$DR, all.trees.join$ND), digits = 2)`, $\lambda_{ND}$ - $\lambda_{BAMM}$: *r*=`r round(cor(all.trees.join$ND, all.trees.join$mean.lambda), digits = 2)`; \href{https://justincally.github.io/SexualSelection_Speciation/#additional_figures_and_tables}{Figure S15}). Given that the calculation of BAMM rates can be affected by the settings of the run and the use of different priors, we compared the estimate of our MCC tree with that of previous published analyses on birds, and found a high correlation (*r*=`r round(cor(Harvey.vs.us$mean.lambda, Harvey.vs.us$Harvey.lambda, use = "pairwise"), digits = 2)`, \href{https://justincally.github.io/SexualSelection_Speciation/#bamm_measures_of_speciation_and_extinction}{Figure S8}, @Harvey_2017). Full details of the BAMM results are presented as supplementary materials.

@@

![Estimates of the effect of individual sexual selection components included in the PPCA (paternal care, sexual size dimorphism and mating system) on three measures of speciation rate ($\lambda_{DR}$, $\lambda_{ND}$ and $\lambda_{BAMM}$). Estimates are presented as density intervals from pgls models on 300 phylogentic trees that used species with available data for these sexual selection measures (*n* = 2,465). The bar under each density ridge is the 95 % Highest Posterior Density Interval. Given that the mating system variable is categorical, model estimates for three polygynous mating system levels are in reference to a striclty monogamous mating system (0% polygyny). \label{SS_components}](Figures/Individual_SS_Estimates.pdf)

@@

**Sexual dimorphism due to competition within sexes contrasts with the drivers of sexual dichromatism.  Plumage dichromatism can evolve as a consequence of female cryptic choice and be related to extra-bond fertilizations, but not necessarily paternal care or mating system [@owens_1998]. It can also arise as a result of selection on the level of crypsis of the sex that cares for offspring [@Dale_2015]. The fact that traits linked with competition (such as size dimorphism) are the ones associated with higher $\lambda_{DR}$ values – rather than sexual dichromatism – supports the general view that antagonistic interactions and sexual conflict can lead to increased diversity [@Bonduriansky_2011; @Hoskin_2010; @tsuji2020sexual]. Moreover, body size is a trait that influences multiple aspects of the physiology and ecology of a species. Differences in body size (as a result of sexual selection) could be linked to changes in diet, vulnerability to predators or environmental tolerance [@Damuth_1993; @Liow_2008; @Bonduriansky_2011], and such differences could ultimately increase the likelihood of divergence between young lineages. In mammals, sexual selection is suggested to have driven the evolution of large body size which in turn has allowed diversification of ecological strategies in the clade, and higher speciation rates [@Mclain_1993; @Bonduriansky_2011].**

@@

In addition to speciation, sexual selection is hypothesised to affect extinction. Using the model-based approach of BAMM, we found no association between the estimated extinction rate and sexual dichromatism, male-biased sexual selection, or our measures of environmental variability. However, these extinction results should not be regarded as definitive because extinction is notoriously difficult to estimate accurately from phylogenies, principally because different combinations of speciation and extinction rates can give rise to similar patterns of diversity [see @Rabosky_Extinction_2016]. Phylogenetic methods such as BAMM allow for speciation and extinction rates to be estimated using moderately-sized phylogenies, although the ability of BAMM to model evolutionary rate shifts and extinction rates is debated [see, @Beaulieu_2015; @Rabosky_Extinction_2016; @Moore_2016; @Rabosky_BAMM_flawed_2017]. Additionally, while several tip-rate estimates exist for speciation rate (e.g., $\lambda_{DR}$ and $\lambda_{ND}$), tip-rate estimates of extinction rate are difficult to obtain without complex Bayesian models which are sensitive to sampling bias [@Davis_2013]. Although extinction rates can be inferred from alternative sources, such as the fossil record [@Martins_2018], direct observation extinction, or IUCN red list status [@Greenberg_2017], each approach has its limitations. Across the passerine bird phylogeny, we found that BAMM often produced homogeneous speciation and extinction rates for smaller clades showing few rate shifts, which might reduce our power to detect small differences in extinction rates among closely-related taxa [@Rabosky_BAMM_flawed_2017; @Rabosky_diversification_2018]. Thus, this methodological constraint likely decreases our ability to accurately measure the correlation between metrics of sexual selection and the probability of extinction.
